---
title: "W9D1 - Uncertainty and confidence intervals"
author: "Nicolas Wittstock"
date: \today
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---
Variance: how representative is the average
Confidence interval: is NOT how confident you are in your estimate to the true value. It is a range of possible estimates for the true value around your estimated value.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Prerequisite
```{r}
# install.packages("stargazer")

library(stargazer)
library(tidyverse)

rm(list = ls())

# data is simulated, no need to load anything

library(ggplot2)

# Set the seed for reproducibility
set.seed(123)

# Generate the population data with 1000 observations
population <- runif(1000, min = 0, max = 4)

```


# Simulation of random sampling

```{r}
# Calculate the true population mean
true_mean <- mean(population)

# Randomly select 3 sets of 100 observations each
sample1 <- sample(population, size = 100)
sample2 <- sample(population, size = 100)
sample3 <- sample(population, size = 100)

# Calculate the sample means
sample1_mean <- mean(sample1)
sample2_mean <- mean(sample2)
sample3_mean <- mean(sample3)

# Calculate the sample standard deviations
sample1_sd <- sd(sample1)
sample2_sd <- sd(sample2)
sample3_sd <- sd(sample3)

# Calculate the standard error of the mean
se1 <- sample1_sd / sqrt(100)
se2 <- sample2_sd / sqrt(100)
se3 <- sample3_sd / sqrt(100)

# Calculate the margin of error for three different critical values

me1 <-  # 90% confidence level
me2 <-   # 95% confidence level
me3 <-  # 99% confidence level

# Calculate the confidence intervals
ci1 <- 
ci2 <- 
ci3 <- 

# Create a data frame for plotting
df <- data.frame(Sample = c("Sample 1 (90%)", "Sample 2 (95%)", "Sample 3 (99%)"),
                 Mean = c(sample1_mean, sample2_mean, sample3_mean),
                 CI_Lower = c(ci1[1], ci2[1], ci3[1]),
                 CI_Upper = c(ci1[2], ci2[2], ci3[2]))
```


# Visualize your estimations using ggplot. Use the geometric shapes `geom_pointrange()` and `geom_hline()` to show the confidence intervals and where the true parameter lies.


```{r}
# Plot the distribution of the sub-samples with confidence intervals

df %>% 
  ggplot(aes(x = Sample, 
             y = Mean,
             ymin = CI_Lower, 
             ymax = CI_Upper)) 

```





