---
title: "W7D1 - Confounding and prediction"
author: "student name"
date: \today
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Prerequisite
```{r}
# install.packages("stargazer")

library(stargazer)
library(tidyverse)

rm(list = ls())

# load data
df <- read_csv(file = "sim.csv")
```


# Regression analysis

Load the dataset sim.csv and plot a scatterplot with y and x variables using ggplot. Fit a linear fit with argument `se=FALSE`.

```{r}
df %>%
  ggplot(aes(x=x, y=y)) +
    geom_point()
    geom_smooth(method="lm") #adds line of best fit

```


Run a regression of y on x and use the `stargazer()` function to report the estimated table. 

**Note**: remember to set the argument type to choose whether to print the results in \LaTeX (for PDF) or *"text"* (for R console). Also, if you want to knit the document in PDF, make sure to activate the code chunk option `results='asis'`.

```{r}
#manually generate a regression model object
m1 <- lm(y ~ x, data = df)

summary(m1)
coef(m1)

stargazer::stargazer(m1, type="text")
```


# Confounding


Create a scatter plot including z in the color aesthetics and a boxplot using ggplot to compare x and y conditional on z.


```{r}
# create a scatterplot with z
df %>%
  ggplot(aes(x=x,y=y, color=factor(z))) +
    geom_point() +
    geom_smooth(method = "lm")


# boxplot of x and y using Z



```

Now, regress y on x and z. Report the regression model of y on both x and z along with the previous model using stargazer.

```{r}
m2 <- lm(y ~ x + z, data = df)

stargazer::stargazer(m1,m2, type = "text") # Bonus: you can also change type to "latex" which will generate a nice table when the file is knitted.

# we estimated a negative relationship with x, but when we include z, x becomes positive (see the plot)
# z is a confounder.

```



# Prediction



To compare the slope of x between the two estimated models, we can compute the **predicted** values of y for a range of hypothetical x values. To do this, we need to create a `tibble()` or `data.frame()` of hypothetical x values from 1 to 12, and then use the `predict()` function. 

For the second model, which includes z as a covariate, we need to create another dataset of hypothetical values that includes a **mean value of z**. This way, the prediction will be based on changes in x while holding z constant at its mean.


```{r}

hyp_x <- tibble(x=1:12)

pred_m1 <- predict()

hyp_x2 <- tibble(x=1:12,
                 z=mean(df$z))

pred_m2 <- predict()

# combine the datasets using cbind()

```




