---
title: "W4D1 lab - Gapminder"
author: "Student name"
date: "1/22/2024"
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

rm(list = ls()) # clean memory

```


```{r}

# Start by installing this package 

install.packages("ggthemes")

# Load both tidyverse and ggthemes

x
library(tidyverse)
library(ggthemes)

```

## Introduction

Let's revisit several functions or concepts we've learnt from lecture, QSS tutorials, and the first assignment 


* `tapply()`

* `ggplot2`

* `quantile()`

* `ifelse()` or `case_when`

* class `factor`  



## Load the gapminder data again

```{r}
gapminder <- read.csv("gapminder.csv")

head(gapminder)
view(gapminder)
```

## `tapply` for group means

Using the `tapply` function, please find:

1. The average `gdpPercap` for each `continent`

2. The average `gdpPercap` for each `year`

3. The average `gdpPercap` for each `continent` over every `year`

```{r}
tapply(gapminder$gdpPercap, gapminder$continent, mean)

tapply(gapminder$gdpPercap, gapminder$year, mean)
# You can combine multiple index variables in tapply using list()
tapply(gapminder$gdpPercap, list(gapminder$continent, gapminder$year), mean)

# Are there any NAs in the dataset? 
# No :D
# To ignore NA values, add na.rm = T or TRUE

```

```{r}

#Alternatively, we can do the same thing in dplyr (which is a package within tidyverse)
gapminder %>% #pipe operator
  group_by(continent, year) %>% 
  summarise(average_gdp = mean(gdpPercap)) %>%
  ggplot(aes(x = year, y = average_gdp, group = continent, color = continent)) + 
  geom_point(size = 0.75) + # empty plot. Use + sign to specify "geometry layer". 
  geom_line() +
  xlab("") +
  ylab("Average GDP (in dollars)") +
  theme_economist() +

# can assign this workflow to an object.

```

## quantile and ifelse

Using `quantile` and `ifelse` or `case_when` function, please create:

1. An object `gdp_qt` that records the quartiles of the `gdpPercap` variable

```{r}
gdp_qt <- quantile(gapminder$gdpPercap)
gdp_qt
```

2. A new variable `poverty` that takes the value of 1 if `gdpPercap` is lower than or equal to the *lower* quartile; 0 otherwise. What is the `sum` of countries in poverty? And their proportion?

```{r}
gapminder$poverty <- ifelse(gapminder$gdpPercap <= gdp_qt[2], 1, 0)

  
  
```


3. A new variable `gdpPercap_cat` that converts `gdpPercap` into four categories: `poor`, `middle`, `wealthy`, and `very wealthy` based on quartiles in `gdp_qt`

```{r}

# You can stack multiple conditions in ifelse
gdpPercap_cat <- ifelse(gapminder$gdpPercap <= gdp_qt[2], "poor",
                        ifelse(gapminder$gdpPercap <=))



```


```{r}
# There is a way to do the same thing in dplyr 


```


4. Use `tapply` to find the mean of `lifeExp` for each income group, based on `gdpPercap_cat`

```{r}

```

5. Do the same thing using `group_by` and `summarise` 

```{r}

```



## Factor

How to inform `R` that `gdpPercap_cat` has an inherent order? 

1. Check out the class of `gdpPercap_cat`

2. Use `factor()` to convert `gdpPercap_cat` into factor, and specify the `levels = c(...)` argument

3. Check out the class of `gdpPercap_cat` again

4. Use `tapply` to find the mean of `lifeExp` for each income group, based on `gdpPercap_cat`

```{r}

```


\pagebreak

## Appendix: R Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```







