---
title: "W4D2 lab - Gapminder, Visualizing Distributions"
author: "Student name"
date: "1/24/2024"
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

rm(list = ls()) # clean memory

```


```{r}

# Start by installing this package 

# install.packages("ggthemes")

# Load both tidyverse and ggthemes

library(tidyverse)
library(ggthemes)

```

## Introduction

Let's revisit several functions or concepts we've learnt from lecture, QSS tutorials, and the first assignment 

* `tapply()`

* `ggplot2`

* `quantile()`

* `ifelse()` or `case_when`


## Load the gapminder data again

```{r}
gapminder <- read.csv("gapminder.csv")

```

## quantile and ifelse

Last time, we created an object that records the quartile values of `gdpPercap`

```{r}
gdp_qt <- quantile(gapminder$gdpPercap)
```

2. Now, we want to create a new variable `gdpPercap_cat` that converts `gdpPercap` into four categories: `poor`, `middle`, `wealthy`, and `very wealthy` based on quartiles in `gdp_qt`

```{r}

# You can stack multiple conditions in ifelse
gapminder$gdpPercap_cat <- ifelse(gapminder$gdpPercap <= gdp_qt[2], "poor", 
                             ifelse(gapminder$gdpPercap > gdp_qt[2] & gapminder$gdpPercap <= gdp_qt[3], "middle", 
                                    ifelse(gapminder$gdpPercap > gdp_qt[3] & gapminder$gdpPercap <= gdp_qt[4], "wealthy",
                                           ifelse(gapminder$gdpPercap > gdp_qt[4], "very wealthy", NA))))
```


```{r}
# There is a way to do the same thing in dplyr 
# And, there is another function to do this, called case_when
gapminder <- gapminder %>%
  mutate(gdpPercap_cat = case_when(gdpPercap < gdp_qt[2] ~ "poor", 
      gdpPercap > gdp_qt[2] & gdpPercap <= gdp_qt[3] ~ "middle", 
      gdpPercap > gdp_qt[3] & gdpPercap <= gdp_qt[4] ~ "wealthy", 
      gdpPercap > gdp_qt[4] ~ "very wealthy"))

```


3. Use `tapply` to find the mean of `lifeExp` for each income group, based on `gdpPercap_cat`

```{r}
#       vector, index, function
tapply(gapminder$lifeExp, gapminder$gdpPercap_cat, mean)
```

4. Do the same thing using `group_by` and `summarise` - and create a new object

```{r}
gapminder %>%
  group_by(gdpPercap_cat) %>%
  summarise(mean = mean(lifeExp))
```



# Visualizing Distributions

We can find means, medians, and quartiles of variables - but is often
more effective to display the distributions of variables visually. 

How do create different graphs in ggplot? 

Recommendation:


https://r-graph-gallery.com/index.html



Histogram
```{r}

#Histograms 

ggplot(gapminder, aes(x = lifeExp)) + 
  geom_histogram(binwidth = 10)

#change the size of the bins using binwidth = 10 within the geom_histogram call


# If we want to plot density, we can use a different geom 

ggplot(gapminder, aes(x = lifeExp)) + 
  geom_density()

# You can also create a histogram based on densityh 

ggplot(gapminder, aes(x = lifeExp)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") +
  theme_wsj()

# notice the changed aesthetic call in geom_histogram -- we tell R that we do not 
# want to plot the counts but rather the proportions 

#Again, there are endless customization options 

ggplot(gapminder, aes(x = lifeExp)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") +
  geom_density(alpha =.2, color = "red") +
  theme_classic() +
  geom_vline(aes(xintercept = median(lifeExp)),
             color = "blue", linetype = "dashed", size = 0.9) +
  labs(title="Life Expectancy Plot", x="Years", y = "Proportion")


```

# Boxplots in ggplot

Let's say we are interested in how life expectancy differs by continent

We could calculate grouped averages (or other grouped statisics)
```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(mean = mean(lifeExp))

```
* We can do something similar visually using a boxplot

```{r}
# Basic box plot

ggplot(gapminder, aes(x= continent, y= lifeExp)) + 
  geom_boxplot()

# This shows us the distribution of lifeExp by continent
# Bar show median, box is 25th and 75th percentile - whiskers show 1.5x IQR 
# The dots further out than the whiskers are typically considered outliers 


# This is a substantively much more useful look at the data 


# Again, we can customize 

# install.packages(c("hrbrthemes", "viridis"))

library(hrbrthemes)
library(viridis)

gapminder %>%
  # for specific year
  filter(year == 2007) %>%
  ggplot(aes(x =continent, y= lifeExp, fill=continent)) +
    geom_boxplot(size = 0.4) +
    scale_fill_viridis(discrete = TRUE, alpha=0.4) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_classic() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Distrbution of Life expectancies by continent") +
    xlab("") + ylab("")

```








