---
title: "W6D1 lab - Graphing Linear Relationships"
author: "Student name"
date: "2/5/2024"
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

rm(list = ls()) # clean memory
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Today's lab

1. Scatterplots 

2. Linear Relationships between two variables 



```{r}
library(tidyverse)
```



## Load the gapminder data again

```{r}
gapminder <- read.csv("gapminder.csv")

head(gapminder)
```


--> We have previously discussed various graphs and different graphing approaches

--> Also, we are interested in finding causal relationships between phenomena

--> If we theorize there to be a causal relationship between A and B, 
    it is typically a good start to graphically assess whether the 
    variables are related. 
    
  The main workhorse for this is the scatterplot. 

  Let's say we want investigate if wealthier people live longer on average:

# Scatterplotting in base r 

```{r}
# We can create a scatterplot of two continuous variables: 

plot(x=gapminder$gdpPercap, y=gapminder$lifeExp)
```

# Customizations 

```{r}
plot(x=gapminder$gdpPercap, y=gapminder$lifeExp,
     xlim=c(0, 90000) , ylim=c(20, 90), 
     pch= 1, 
     cex=.2, 
     col="#69b3a2",
     xlab="GDP Per Capita", ylab="Life Expectancy",
     main="Base R scatterplot"
     )
```

# What seems to be going on here? What kind of relationship is there? 

# Let's manipulate some of these values and see what happens 





# Scatterplots in ggplot 

# Basic syntax

```{r}
gapminder %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```


# What seems to be the relationship between GDP PC and average life expectancy? 

--> Some positive, not necessarily linear relationship

# What is important to notice about the gdpPercap variable? 

```{r}
summary(gapminder$gdpPercap)
```

--> 

```{r}
hist(gapminder$gdpPercap)
```

# It looks like there are a lot of country-year values that are quite low, 
# and just very few that are higher than 20k/year

--> Global wealth inequality 

--> The average Singaporean person in 2007 is dramatically more wealthy in real
    monetary terms than the average Ethiopian in 1952. 
    
# Let's identify the smallest / highest values

```{r}



```

# What are the smallest / highest values for each year and continent? 

# And, what countries are these? 

```{r}
gapminder %>%
  group_by(year) %>%
  summarise(min = min(gdpPercap), 
            country = country[which.min(gdpPercap)])

gapminder %>%
  group_by(year) %>%
  summarise(max = max(gdpPercap), 
            country = country[which.max(gdpPercap)])

```

    

    
# What can we do to study how average personal wealth and life expectancy are 
# related? 





# We limit our analysis to one year

```{r}
gapminder %>%
  filter(year == 1967) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()  + 
  theme_classic()
```


# *Or one continent 

```{r}
#This pools all country-year observations for countries within Asia

gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  theme_classic()
```

# Or, we can look at one specific country and look at measurements of wealth 
# and lifeexpectancy at different points in time 

```{r}
gapminder %>%
  filter(country == "India") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  theme_classic()
```


# Ok, let's pool all observations again  

```{r}
gapminder %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(size = .5) + 
  theme_classic()
```

--> When we look at all of our country-year pairs, it becomes apparent that 
    there seem to be a lot of observations with low gdpPercap 
    
    
    
    
    
## Log transforming gdpPercap 

```{r}
gapminder$log_gdp <- log(gapminder$gdpPercap)

# A log transformation retains the scale, but effectively compresses the distance
# between observations. 
```

# Now, let's look at another scatterplot 

```{r}
gapminder %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() + 
  theme_classic() 
  
```

--> How would you now describe the relationship between GDP and life expectancy?


--> What issue do you notice with our X axis? 



```{r}
gapminder %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() + 
  theme_classic() +
  scale_x_continuous(
    breaks = log(c(300, 3000, 30000)),  # I create log-transformed breaks
    labels = c("300 USD", "3,000 USD", "30,000 USD")  # Labels in real USD
  ) +
  labs(x = "GDP per Capita (USD)")
```



Conceptually, why might GDPpc and life expectancy be related? 

# Do you think that this relationship might differ by continent? 

--> Why might that be so? 

# Let's investigate visually 

```{r}
gapminder %>%
  filter(continent == "Africa") %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() +
  theme_classic()

gapminder %>%
  filter(continent == "Americas") %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() +
  theme_classic()
```

# We can do even better in ggplot 

```{r}
gapminder %>%
  ggplot(aes(x = log_gdp, y = lifeExp, group = continent, color = continent)) +
  geom_point() +
  theme_classic() +
  theme(legend.position = "bottom") +
  scale_x_continuous(
    breaks = log(c(300, 3000, 30000)),  
    labels = c("300 USD", "3,000 USD", "30,000 USD")  
  ) +
  labs(x = "GDP per Capita (USD)")
```

# Now, let's get a bit more fancy 

```{r}

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = log_gdp, y = lifeExp, group = continent, color = continent,
             shape = continent)) +
  geom_point(alpha = .3) +
  theme_classic() +
  theme(legend.position = "bottom") +
  geom_jitter() +
  geom_rug(alpha = 1/4, position = "jitter") +
  scale_x_continuous(
    breaks = log(c(300, 3000, 30000)),  
    labels = c("300 USD", "3,000 USD", "30,000 USD") 
  ) +
  labs(x = "GDP per Capita (USD)")


```


# At minimum, we can see that African countries look to be in the lower left 
# quadrant - on average; and Europen countries in blue are in the upper right

# Let's look at that more closely

```{r}
gapminder %>%
  filter(year == 2007, continent == "Africa") %>%
  ggplot(aes(x = log_gdp, y = lifeExp, group = continent, color = continent)) +
  geom_point(alpha = .3) +
  theme_classic() +
  theme(legend.position = "none") +
  geom_jitter() +
  geom_rug(alpha = 1/2, position = "jitter") 
```

```{r}
gapminder %>%
  filter(year == 2007, continent == "Europe") %>%
  ggplot(aes(x = log_gdp, y = lifeExp, color = "red")) +
  geom_point(alpha = .3) +
  theme_classic() +
  theme(legend.position = "none") +
  geom_jitter() +
  geom_rug(alpha = 1/2, position = "jitter") +
  geom_text_repel( aes(label = country) )
```

# Does anyone notice a pattern across European countries? 



# Let's look at some specific countries 

```{r}

library(ggrepel)
library(MASS)
library(RColorBrewer) 

# Here, I am specifying a specific color palette -- instead of using ggplot's 
# defaults. 

colors <- brewer.pal(n = 5, "Set1") 
# Here, I am choosing specific colors to use below

red <- colors[1] 
blue <- colors[2] 
green <- colors[3]
orange <- colors[4]
purple <- colors[5]


gapminder %>%
  filter(year == 2007,
         country %in% c("Germany", "Japan", "United States", "France", "Mexico",
                        "China", "Kuwait", "India", "Brazil", "Tunisia", "Egypt",
                        "Saudi Arabia", "Iran", "Australia")) %>%
  ggplot(aes(x = log_gdp, y = lifeExp, group = continent, color = continent)) +
  geom_point(alpha = .3) +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_blank()) +
  geom_rug(alpha = 1/2, position = "jitter") +
  geom_text_repel( aes(label = country) ) + 
  ggtitle("Life Expectancy and GDP pc in 2007") + 
  xlab("GDP per Capita (USD)") + 
  ylab("Life Expectancy in years") +
    scale_color_manual( 
    values = c( "Europe" = green, 
                "Asia" = red, 
                "Americas" = blue,
                "Africa" = orange,
                "Oceania" = purple
                )) +
   scale_x_continuous(
    breaks = log(c(3000, 10000, 40000)),  
    labels = c("3000 USD", "10,000 USD", "40,000 USD") 
  )

```


# Ok, let's take a step back 

# There is a clear positive relationship between wealth and life expectancy

# But, it looks like this relationship is not identical across the world

```{r}

gapminder %>%
  filter(!continent == "Africa") %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() + 
  theme_classic()


gapminder %>%
  filter(continent == "Africa") %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() + 
  theme_classic()

```

# What is the correlation between these variables? 

```{r}
cor(gapminder$gdpPercap, gapminder$lifeExp)

cor(gapminder$log_gdp, gapminder$lifeExp)
```

# Is this correlation different for each continent? 
```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(correlation = cor(gdpPercap, lifeExp))
```

# There are dramatic differences in how strongly these variables are 
# linearly related across continents! 

--> It appears that whatever the causal effect of wealth on life expectancy might 
    be -- it is different across the world
    
    
# Let's look at Asia 

```{r}
gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  theme_classic()
```

# And Oceania

```{r}
gapminder %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() + 
  theme_classic()
```

# on the same plot 

```{r}
gapminder %>%
  filter(continent %in% c("Oceania", "Asia")) %>%
  ggplot(aes(x = log_gdp, y = lifeExp, color = continent)) +
  geom_point() + 
  theme_classic()
```

# ggplot has the option of fitting a linear model to the data (linear regression)

```{r}

gapminder %>%
  ggplot(aes(x = log_gdp, y = lifeExp)) +
  geom_point() + 
  theme_classic() +
  geom_smooth(method = "lm")

# --> This depicts a linear regression line that is fitted to the observations
```

# Does the linear model differ for each continent? It really should as we have seen

```{r}
gapminder %>%
  ggplot(aes(x = log_gdp, y = lifeExp, group = continent, color = continent)) +
  geom_point(size = .3) + 
  theme_classic() +
  geom_smooth(method = "lm")
```

