---
title: "W6D2 - Regression and Confounding"
author: "student name"
date: /today
output:
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Prerequisite
```{r}
library(tidyverse)

rm(list = ls())

# load data
data(mtcars)
```


# Regression analysis

Perform a regression of mpg ($Y$) on hp ($X$) and report the estimated coefficient using the function `lm()` and `summary()`. 

```{r}
fit <- lm(mpg ~ hp, data = mtcars)
summary(fit)$r.squared
```


Now, Visualize the relationship between mpg and hp using ggplot and fit a linear regression.

```{r}
mtcars %>%
  ggplot(aes(x=hp, y=mpg)) +
  geom_point() +
  geom_smooth(method = "lm")
```



# Confounding


Create a new plot by adding a color aesthetic that maps the `cyl` variable. Ensure that `cyl` is of class factor so that ggplot automatically groups each level. 

```{r}
mtcars %>%
  ggplot(aes(x=hp, y=mpg, group=cyl)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic()

```


Then, perform a regression of `mpg` on `hp` and `cyl`. Compare the results with the previous model to identify any changes.

```{r}
fit_2 <- lm(mpg ~ hp + cyl, data=mtcars)
coef(fit_2) #we see that cyl has a huge effect on mpg.
#What is the effect if we hold the confounding constant?

```

