---
title: "W2D2_gapminder"
author: "Ariel Lin"
date: \today
output: pdf_document
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
#include=FALSE tells R not to include when knitting
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls()) # clean memory from the global environment
```

## Prerequisite

Please install the following package if you haven't by un-commenting out the first line (`install.packages(...)`). The keyboard shortcut to comment out a line is `Ctrl` + `Shift` + `C`, or for macOS, `Command` + `Shift` + `C`

```{r}
# install.packages("tidyverse") ## You just have to run it once
install.packages("tidyverse") ## You just have to run it once
library(tidyverse)
library(dplyr)              ## load the dplyr package
```

## Load the data 

Load the gapminder data using `read.csv()`. Make sure that the file is in the right directory.
```{r}
setwd('/Users/Harry/Documents/Ariel/~College/~first year/WIN 24/STAT 321/W2D2')
getwd()
read.csv("gapminder.csv")
# getwd()


```

## Inspect the data

Please find out: 
  1) The class of the `data` object
  2) Snippet of the top and bottom rows
  3) Its numbers of rows and columns
  4) The names of columns/variables
  5) Summary of all columns/variables
  6) Mean and standard deviation of the `gdpPercap` variable: 

```{r}
data<-read.csv('gapminder.csv')

# 1 find the class of the data object
class(data) #data.frame

# 2 Snippet of the top and bottom rows
data[1,] #top row
head(data) #shows first 6 rows
# nrow(data)
data[1704, ] #bottom row
tail(data) #shows last bottom 6 rows

# 3 number of rows and columns - rows are always mentioned first
dim(data)
nrow(data)
ncol(data)

# 4 names of columns/variables
colnames(data)
names(data)

# 5 summary of all columns/variables
summary(data)

# 6 Mean and standard deviation of gdpPercap variable
mean(data$gdpPercap)
sd(data$gdpPercap)

```

When did the data start and end? 
```{r}
# ask what this question is asking for, are we asking for years?
data$year[1]
data$year[1704]
```

How many unique countries there are in the data?
```{r}
length(unique(data$country))
# there are 142 unique countries in the data
```

## Subset the data

Subset the data for `Afghanistan` since 1980. 

```{r}
subset(data, (country == "Afghanistan" & year>=1980))
```


## Transformation

Create the following new columns/variables: 
  1. Create `gdpTotal` by multiplying `gdpPercap` and `pop`
  2. Create `gdpMil` that transform `gdpTotal` in millions, and round it up to one decimal place
  3. Create `gdpMil.prev` that records the `gdpMil` of the previous year (with the help of `lag()`)
  4. Create `gdpChange` that computes the change in `gdpMil` (presented in percentage point and round up to one decimal place)
  5. Plot `gdpChange` over year for Afghanistan
```{r}
# 1
gdpTotal<-c(data$gdpPercap * data$pop)
# gdpTotal
gdpMil<-round(gdpTotal/(10^6), digits=1) #????????????????????????
# gdpMil.prev
gdpMil.prev<-lag(gdpMil)

```



## Plot the data

Plot the following variables using the `plot()` function: 
  1. Plot life expectancy over time for Afghanistan data
  2. Plot gdp per capita over time
  3. Plot life expectancy against gdp per capita

If you don't know what `plot()` function is doing, run `?plot` to access the help page 
```{r}
# ?plot
dataAfghanistan<-subset(data, (country == "Afghanistan"))
# summary(dataAfghanistan)
plot(dataAfghanistan$lifeExp, dataAfghanistan$year, xlab="Life Expectancy", ylab="Year")
title(main="Life Expectancy Over Time for Afghanistan")
#
plot(dataAfghanistan$gdpPercap, dataAfghanistan$year, xlab="GDP per Capita", ylab="Year")
title(main="GDP per Capita Over Tie for Afghanistan")
```


## Sorting

Find out:
  1. the top 10 country-year observations with highest `gdpPercap`
  2. the bottom 10 country-year observations with lowest `gdpPercap`

```{r}
head(data$gdpPercap)
summary(data$gdpPercap)
```





