---
title: "321 Final Project"
author: "Ariel Lin and Naomi Le Mouel"
date: "`r Sys.Date()`"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
output:
  pdf_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE, float = FALSE)


# load libraries
library(tinytex)
library(questionr)
library(broom)
library(stargazer)
library(tidyverse)
library(data.table)
library(xtable)

rm(list = ls())


# load data
socDem_raw <- read_csv("socialDemo.csv")
```

```{r, include=FALSE}
#cleaning data
vars_of_interest <- c("SATISFIED_DEMOCRACY_W105","SNSUSE_W105",
                      "POST_ISSUES_W105","SOCIAL_DEMOCRACY_W105",
                      "F_AGECAT", "F_IDEO","F_RACECMB",
                      "SMIMPACT_INFORMED_W105","SMIMPACT_DISCOURSE_W105",
                      "SMIMPACT_MANIPULATE_W105","SMIMPACT_DIVIDED_W105")

socDem <- socDem_raw[,(names(socDem_raw) %in% vars_of_interest)]

#quantifying data
#variables of interest
  #General 
socDem$social_demo <- (
  ifelse(socDem$SOCIAL_DEMOCRACY_W105 == "More of a GOOD THING for democracy in the U.S.", 1,
  ifelse(socDem$SOCIAL_DEMOCRACY_W105 == "More of a BAD THING for democracy in the U.S.", 0, 
  NA)))

  #Specifics: positive
socDem$smp_informed <- (ifelse(socDem$SMIMPACT_INFORMED_W105 == "More", 1,
                          ifelse(socDem$SMIMPACT_INFORMED_W105 == "Less", 2,
                          ifelse(socDem$SMIMPACT_INFORMED_W105 == "Not had much impact", 3,
                          NA))))

socDem$smp_civil <- (ifelse(socDem$SMIMPACT_DISCOURSE_W105 == "More", 1,
                          ifelse(socDem$SMIMPACT_DISCOURSE_W105 == "Less", 2,
                          ifelse(socDem$SMIMPACT_DISCOURSE_W105 == "Not had much impact", 3, 
                          NA))))
  #Specifics: negative
socDem$smn_manip <- (ifelse(socDem$SMIMPACT_MANIPULATE_W105 == "More", 1,
                          ifelse(socDem$SMIMPACT_MANIPULATE_W105 == "Less", 2,
                          ifelse(socDem$SMIMPACT_MANIPULATE_W105 == "Not had much impact", 3, 
                          NA))))

socDem$smn_divided <- (ifelse(socDem$SMIMPACT_DIVIDED_W105 == "More", 1,
                          ifelse(socDem$SMIMPACT_DIVIDED_W105 == "Less", 2,
                          ifelse(socDem$SMIMPACT_DIVIDED_W105 == "Not had much impact", 3, 
                          NA))))

#Dependent
socDem$satisfied <- (ifelse(socDem$SATISFIED_DEMOCRACY_W105 == "Very satisfied", 4,
                          ifelse(socDem$SATISFIED_DEMOCRACY_W105 == "Somewhat satisfied", 3,
                          ifelse(socDem$SATISFIED_DEMOCRACY_W105 == "Not too satisfied", 2, 
                          ifelse(socDem$SATISFIED_DEMOCRACY_W105 == "Not at all satisfied", 1, 
                          NA)))))

#Systematically different groups
socDem$socialuse <- (ifelse(socDem$SNSUSE_W105 == "Yes, I use social media sites", 1,
                            ifelse(socDem$SNSUSE_W105 == "No, I do not use social media sites", 0,
                            NA)))


#Other variables (possible heterogeneous treatment effects)
socDem$ideology <- (ifelse(socDem$F_IDEO == "Liberal" | socDem$F_IDEO == "Very liberal" , 1,
                      ifelse(socDem$F_IDEO == "Moderate", 2,
                      ifelse(socDem$F_IDEO == "Conservative" | socDem$F_IDEO == "Very conservative", 3,
                      NA))))
  #Is ideology related to satisfaction with democracy? 

socDem$posting <- (ifelse(socDem$POST_ISSUES_W105 == "Often", 1,
                          ifelse(socDem$POST_ISSUES_W105 == "Sometimes", 2, 
                          ifelse(socDem$POST_ISSUES_W105 == "Rarely", 3,
                          ifelse(socDem$POST_ISSUES_W105 == "Never", 4, 
                          NA)))))
  #Do people who post social and political issues on social media believe social media has a "good"
  #impact on democracy? (Believing they can use the platform to educate others on issues etc.)

socDem$agecat <- (ifelse(socDem$F_AGECAT == "18-29", 1,
                         ifelse(socDem$F_AGECAT == "30-49", 2,
                         ifelse(socDem$F_AGECAT == "50-64", 3,
                         ifelse(socDem$F_AGECAT == "65+", 4,
                         NA)))))
  #Do older people perceive social media's impacts on democracy more negatively than younger people?

socDem$race <- (ifelse(socDem$F_RACECMB == "Mixed Race", 1,
                       ifelse(socDem$F_RACECMB == "Asian or Asian-American", 2,
                       ifelse(socDem$F_RACECMB == "Or some other race", 3,
                       ifelse(socDem$F_RACECMB == "Black or African-American", 4,
                       ifelse(socDem$F_RACECMB == "White", 5,
                       NA))))))
  #Are non-White minorities less satisfied with the U.S. democracy?


#only quantitative variables
dat_socDem <- socDem[,c("posting","social_demo","satisfied","ideology","socialuse","agecat","race",
                        "smn_manip","smn_divided","smp_informed","smp_civil")]

#looking at the data
summary(dat_socDem)
freq.na(dat_socDem) # the relative and absolute NAs for each variable

```


# Introduction


Social media has a powerful presence in American lives. According to data from Statista, there are over 300 million social media users in the United States. Media platforms are used for many purposes, including entertainment, socialization, business, education, and news. Around half of the US population gets their news information from social media, according to a 2021 study conducted by the Pew Research Center. With so many media influencers and users, there is a lot of room for misinformation, fake news, propaganda, and online discourse, which can often concern political topics such as presidential elections. 

In this report, we examine the potential relationship between Americans’ satisfaction with U.S. democracy and their opinions on social media’s impact on democracy. The data used in this report is from the Pew Research Center’s American Trends Panel (ATP) Wave 105, which surveyed 3,581 U.S. adults from March 21 to 27, 2022, through random sampling of residential addresses and weighted to represent the U.S. population by gender, race, and other demographics.  


# Theory and Hypotheses


Using the available ATP Wave 105 data, we want to answer this question: what effect has social media had on democracy in the eyes of adults in the United States? In other words, how is American satisfaction with U.S. democracy related to their opinions on social media's impact on democracy? 
Our report draws from a broader Pew Research Center report by Wike et al. published in December 2022 that observes the trend in attitudes toward social media and its effects on democracy in several countries across the globe. The study used the ATP Wave 105 data for the U.S. and randomized phone surveys for non-U.S. countries and found that social media was perceived as “mostly good” by most countries in general, except the United States. Attitudes toward social media’s effect on democracy are further explored more specifically through survey questions asking about social media’s impact on positive outcomes, such as better informing people, and negative outcomes, such as manipulating people.

Here, we focus on the United States and make the causal assumption that low American satisfaction with democracy in the US should have a positive correlation with the belief that social media harms democracy (and vice versa). We reason that the people who believe social media is “bad” for democracy are also unsatisfied with the US democracy because social media contributed to the unsatisfactory democracy.

Based on our assumption, two hypotheses are considered:

- $H_0$ (null hypothesis) - American adult’s belief that social media is “bad” or “good” for democracy does not affect satisfaction/dissatisfaction with democracy.

- $H_1$ (alternative hypothesis) - American adult’s belief that social media is “bad” or “good” for democracy does affect satisfaction/dissatisfaction with democracy.

We would also like to consider a few possible heterogeneous effects: 

* Posting social/political issues on social media: Do people who post social and political issues on social media believe social media has a "good" impact on democracy? (Believing they can use the platform to educate others on issues etc.)
  
* Ideology: are liberal respondents more likely to be dissatisfied with the U.S. democracy? Are moderates more likely to be satisfied?

* Age: do older people perceive social media's impacts on democracy more negatively than younger people?

* Race: Are non-White minorities less satisfied with the U.S. democracy?


# Data and Visual Analysis

## Data

The ATP Wave 105 dataset conducted by the Pew Research Center contains 3,581 observations representing survey questionnaire responses of individual U.S. adults and 159 variables corresponding to each question on the survey. Our study is observational since we are looking at relationships between responses to relevant questions on the ATP Wave 105 questionnaire.

We selected the following questions and variables to focus on for our project:

`SATISFIED_DEMOCRACY_w105`: How satisfied are you with the way democracy is working in the U.S?

1) Very satisfied

2) Somewhat satisfied

3) Not too satisfied

4) Not at all satisfied

`F_PARTY_FINAL`: As of today do you lean more to

1) Republican

2) Democrat

3) Independent

4) Something else

`SNSUSE`: Do you use social media sites like Facebook, Twitter, or Instagram?

1) Yes, I use social media sites

2) No, I do not use social media sites

`SOCIAL_DEMOCRACY_w150`: Overall, when you consider all the advantages and disadvantages of social media, would you say social media has been

1) More of a GOOD THING for democracy in the U.S.

2) More of a BAD THING for democracy in the U.S.

`SMIMPACT_BATT`: In general, do you think access to the internet and social media has made people more or less

  `SMIMPACT_INFORMED_w105`: Informed about current events in our country?
  
  `SMIMPACT_DIVIDED_w105`: Divided in their political opinions?
  
  `SMIMPACT_MANIPULATE_w105`: Easy to manipulate with false information and rumors?
  
  `SMIMPACT_DISCOURSE_w105`: Civil in the way they talk about politics?
  
  1) More
  
  2) Less
  
  3) Not had much impact

We also included variables from the dataset that were not associated with any particular question in the survey:

`F_AGECAT`: age category of the respondents

1) 65+

2) 50-64

3) 30-49

4) 18-29

`F_IDEO`: ideology of the respondents

1) Very conservative

2) Conservative

3) Liberal

4) Moderate

5) Very liberal

`F_RACECMB`: race the respondent identifies with

1) White

2) Asian or Asian-American

3) Black or African-American

4) Mixed Race

5) Or some other race


The original response data ATP Wave 105 dataset was qualitative, i.e., stored as words and phrases. To better observe relationships between the variables, we created new variables that quantified the data by assigning numerical values representing the responses, which we stored into a new, smaller dataset of 11 quantitative variables instead of 159. We also created a dataset with 22 variables that included the original qualitative responses for the variables. A summary of the variables of our quantitative modified dataset is shown below.

-------------------------    ---------------------------------------------------------------
Name                         Description

-------------------------    ---------------------------------------------------------------
`satisfied`                  Satisfaction of respondents with Democracy in the US on a scale 
                             of 1 (not at all satisfied) to 4 (very satisfied)


`socialuse`                  Whether or not respondents use social media sites (yes, 1 or no,
                             0)


`social_demo`                Whether respondents think social media is more of a GOOD (1) 
                             thing or a BAD (0) thing for democracy 


`posting`                    Whether or not respondents posted about political or social 
                             issues on social media on a scale of 1(often) to 4(never)
          
`smp_informed`               If respondents think access to the internet and social media 
                             makes people more or less informed about current events in our 
                             country (more, 1, less, 2, not much impact, 3)
            
`smn_manip`                  If respondents think access to the internet and social media 
                             makes people more or less manipulated with false information
                             and rumors (more, 1, less, 2, not much impact, 3)
                 
`smn_divided`                If respondents think access to the internet and social media 
                             makes people more or less divided in their political opinions (more, 1, less, 2, not much impact, 3)
`smp_civil`                  If respondents think access to the internet and social media 
                             makes people more or less civil in the way they talk about
                             politics (more, 1, less, 2, not much impact, 3)
                        
`ideology`                   The ideology of the respondents on a scale of 1 (liberal) to 3 (conservative)

`agecat`                     The age category the respondent falls under on a scale of 1 to 4 (18-29, 30-49, 50-64,
                             65+)

`race`                       The race of the respondent (mixed race 1, Asian or Asian american 2, some other race 3, Black or African American 4, white 5)
--------------------------------------------------------------------------------
We omitted missing values when conducting our the majority of our data analyses, since most variables have only small proportion of missing values (all under 5 percent). There are 2426 observations that do not contain missing values. The `posting` variable is the only exception, which we assume may be due to some respondents being uncomfortable disclosing whether or not they post political and social issues on social media. (*See Table 1*)

```{r, echo=FALSE, results='asis',fig.show='hold',out.width='70%'}
prop_na_table <- as.data.frame(tibble(Variable = names(dat_socDem), Num_NAs = colSums(is.na(dat_socDem)), 
       Prop_NAs = (colSums(is.na(dat_socDem))/nrow(dat_socDem))))

stargazer(prop_na_table,
          header=FALSE,
          summary=FALSE,
          rownames = FALSE,
          type="latex",
          title="Proportion of Missing Values"
          )

```

The mean of 0.33 for the `social_demo` variable suggests that a third of U.S. adults think social media is “good” for democracy and the majority think social media is “bad” for democracy. Most U.S. adults are also not very satisfied with the U.S. democracy, based on the low mean of 2.23 for `satisfied`. The averages of the specific effects (`smp_inform`, `smp_divided`, `smn_manip`) of social media on democracy are fairly similar, implying that for both positive and negative effects, U.S. adults generally think social media has a “more of an impact” on those effects (manipulation, political division, information) and “less of an impact” on civil discussions about politics (`smp_civil`). The majority of respondents in the dataset are between 30 and 64 years of age, white, use social media, have a moderate ideology, and rarely post about social and political issues on social media. (*See Table 2*)

```{r, results='asis', echo=FALSE, message=FALSE}
# Display well-formatted table using stargazer

stargazer(as.data.frame(dat_socDem), # input the object "as.data.frame" for descriptives
          summary.stat = c("mean","sd","min","median","max"),
          type="latex",
          header = FALSE,
          title = "Descriptive statistics",
          digits = 2, table.placement = "!h"
          )

```

Based on our correlation matrix, the variable `social_demo` exhibits a positive correlation of 0.19 with the variable `satisfied`. Additionally, `posting` demonstrates a small positive correlation of 0.06 with satisfaction with the U.S. democracy, while `ideology` displays a negative correlation of -0.20, suggesting that more respondents with conservative ideologies are more likely to be less satisfied with the U.S. democracy. (*See Table 3*)

```{r, results='asis', echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}

# Display a correlation matrix

# create an object with the correlation matrix
dat_socDem_noNA <- na.omit(dat_socDem)

dat_cor <- cor(dat_socDem_noNA)

# Now we need to format the correlation matrix, start rounding the digits

dat_cor <- round(dat_cor, digits = 2)

# display the correlation matrix in a formmatted table using stargazer

stargazer(dat_cor, 
          type="latex",
          title = "Correlation matrix",
          font.size = "small",
          column.sep.width = "-10pt",
          header = FALSE, 
          digits = 2, table.placement = "!h")

```

## Exploratory Data Analysis

We begin by observing the distribution of U.S. adults' satisfaction with U.S. democracy.

For the entire dataset, including the small number of respondents who do not use social media, the percentage for each response of the total number of respondents is displayed in Table 4. The vast majority of the responses are around "2" and "3", which reflects that most respondents are either "Not too satisfied" (41.7%) or only "Somewhat satisfied" (34.5%) with the U.S. democracy. Note: the last row represents the percentage of missing values, which are under 1 percent for both tables below.

```{r, echo=FALSE, results='asis', float=FALSE}
#subset data so all respondents use social media
usesm_socDem <- subset(dat_socDem, dat_socDem$socialuse > 0)

all <- as.data.table(dat_socDem$satisfied, useNA = 'always')
sm_use <- as.data.table(usesm_socDem$satisfied, useNA = 'always')

#tables
table_all <- all[, .(Percentage = .N / nrow(all) * 100), by = dat_socDem$satisfied]
names(table_all) <- c("Satisfaction","% of Respondents")

stargazer(as.data.frame(table_all), # input the object "as.data.frame" for descriptives
          summary=FALSE,
          type="latex",
          rownames=FALSE,
          header = FALSE,
          title = "Distribution of Satisfaction with U.S. Democracy",
          digits = 2
          )

```

Subsetting the data for only those respondents who *do* use social media, we still see a similar trend, with a slightly higher percentage for "Not too satisfied" (about 2 percentage points higher) with U.S. democracy.

```{r, echo=FALSE, results='asis', float=FALSE}
table_sm <- sm_use[, .(Percentage = .N / nrow(sm_use) * 100), by = usesm_socDem$satisfied]
names(table_sm) <- c("Satisfaction","% of Respondents")

stargazer(as.data.frame(table_sm), # input the object "as.data.frame" for descriptives
          summary=FALSE,
          type="latex",
          rownames=FALSE,
          header = FALSE,
          title = "Distribution of Satisfaction with U.S. Democracy among Social Media Users",
          digits = 2
          )
```

Plotting the percentages by category of satisfaction for all respondents and for only respondents that use social media, we again see that the distributions are very similar, with the majority of responses in the "Not too satisfied" with U.S. democracy category.

```{r, echo=FALSE, fig.show='hold', out.width='50%'}
#Use vs Don't use social media
table_sm <- sm_use[, .(Percentage = .N / nrow(sm_use) * 100), by = usesm_socDem$satisfied]

barplot(Percentage ~ usesm_socDem, table_sm,
        main = "Percentage of Social Media Users' \n Satisfaction with American Democracy",
        xlab = "Categories of Satisfaction", names = c("Not at all", "Not too Satisfied", "Somewhat", "Very         Satisfied"),
        ylab = "Percentage of Respondents", ylim = c(0, 50), 
        col="#93affa")

table_all <- all[, .(Percentage = .N / nrow(all) * 100), by = dat_socDem$satisfied]

barplot(Percentage ~ dat_socDem, table_all,
        main = "Percentage of All Respondents' \n Satisfaction with American Democracy",
        xlab = "Categories of Satisfaction", names = c("Not at all", "Not too Satisfied", "Somewhat", "Very         Satisfied"),
        ylab = "Percentage of Respondents", ylim = c(0, 50),
        col="#93affa")
```

Now we compare the satisfaction with U.S. democracy of respondents who believe social media is more of a bad thing for democracy with that of respondents who believe social media is more of a good thing. We subset the data into these two groups and observe the trends in the percentage of responses for each satisfaction category.

Based on Table 7, the majority of responses (43.61%) among U.S. adults that believe social media is more of a *bad* thing for democracy are "2", meaning they are "Not too satisfied" with U.S. democracy.

```{r, echo=FALSE, results='asis', float=FALSE}
#Subset data by GOOD or BAD
B <- subset(dat_socDem, dat_socDem$social_demo == 0)
G <- subset(dat_socDem, dat_socDem$social_demo == 1)

BAD <- as.data.table(B$satisfied, useNA = 'always')
GOOD <- as.data.table(G$satisfied, useNA = 'always')

#tables
table_BAD <- BAD[, .(Percentage = .N / nrow(BAD) * 100), by = B$satisfied]
names(table_BAD) <- c("Satisfaction","% of Respondents")

stargazer(table_BAD, # input the object "as.data.frame" for descriptives
          summary=FALSE,
          type="latex",
          rownames=FALSE,
          header = FALSE,
          title = "Satisfaction with U.S. Democracy among Respondents Believing Social Media is Bad for Democracy",
          digits = 2
          )

```

In contrast, the majority of responses (43.71%) among U.S. adults that believe social media is more of a *good* thing for democracy are "3", meaning they are "somewhat satisfied" with U.S. democracy.

```{r, echo=FALSE, results='asis', float=FALSE}
table_GOOD <- GOOD[, .(Percentage = .N / nrow(GOOD) * 100), by = G$satisfied]
names(table_GOOD) <- c("Satisfaction","% of Respondents")

stargazer(table_GOOD, # input the object "as.data.frame" for descriptives
          summary=FALSE,
          type="latex",
          rownames=FALSE,
          header = FALSE,
          title = "Satisfaction with U.S. Democracy among Respondents Believing Social Media is Good for Democracy",
          digits = 2
          )
```

Comparing the percentage of respondents who responded "Not too satisfied" for the two tables, the percentage of respondents who believe social media is more of a good thing for democracy is 5.35 percentage points lower than that of respondents who believe social media is more of a bad thing for democracy. 

```{r, include=FALSE}
#Difference in percentages for responding 2 (Not too satisfied)
#responded 2 GOOD
table_GOOD[,2][1] - table_BAD[,2][2]
#  -5.348534
```

The bar plots help illustrate our observations. Comparing the two bar plots, we see that the distribution for respondents believing social media is bad for democracy is slightly more skewed right, while the distribution for respondents believing social media is good for democracy is slightly more skewed to the left. The skewed distributions indicate that those who believe social media is good for democracy are more likely to be satisfied with U.S. democracy and those who believe social media is bad for democracy are less likely satisfied with the U.S. democracy.

```{r, echo=FALSE, fig.show='hold', out.width='50%'}

table_GOOD <- GOOD[, .(Percentage = .N / nrow(GOOD) * 100), by = G$satisfied]
table_BAD <- BAD[, .(Percentage = .N / nrow(BAD) * 100), by = B$satisfied]

#Thought SM was Good vs Bad for US democracy
barplot(Percentage ~ B, table_BAD,
        main = "Percentage of Respondents who Think Social Media \n has a Negative Effect on American Democracy and \n their Satisfaction with American Democracy",
        xlab = "Categories of Satisfaction", names = c("Not at all", "Not too Satisfied", "Somewhat", "Very         Satisfied"),
        ylab = "Percentage of Respondents", ylim = c(0, 50),
        col="#e08b8b")

barplot(Percentage ~ G, table_GOOD,
        main = "Percentage of Respondents who Think Social Media \n has a Positive Effect on American Democracy and \n their Satisfaction with American Democracy",
        xlab = "Categories of Satisfaction", names = c("Not at all", "Not too Satisfied", "Somewhat", "Very         Satisfied"),
        ylab = "Percentage of Respondents", ylim = c(0, 50),
        col="#88c758")
```

For more specific impacts of social media, the Pew Research study considers the variables `smp_informed` and `smp_civil` good impacts of social media and `smn_manip` and `smn_divided` bad impacts of social media on democracy (see the table on page 4 for descriptions of the variables). We presume that respondents who believe social media is good for democracy will more likely believe that social media has "more of an impact" (coded as a 1 in our quantitative variable) on the "good" variables and less of an impact or no impact (2 or 3) on the "bad" variables. The opposite might be true for respondents believing social media has an overall bad impact on democracy. 

```{r, results='asis', echo=FALSE, fig.show='hold', out.width='50%'}
mean_GOOD_impacts <- data.frame(
  Impact = c("smp_civil","smp_informed","smn_manip","smn_divided"),
  Mean = c(
    mean(G$smp_civil, na.rm=TRUE),
    mean(G$smp_informed,na.rm=TRUE),
    mean(G$smn_manip,na.rm=TRUE),
    mean(G$smn_divided,na.rm=TRUE)
  )
)

stargazer(mean_GOOD_impacts, # input the object "as.data.frame" for descriptives
          summary=FALSE,
          type="latex",
          rownames=FALSE,
          header = FALSE,
          title = "Social Media's Specific Impacts for Respondents Believing Social Media is Good for Democracy",
          digits = 2)

mean_BAD_impacts <- data.frame(
  Impact = c("smp_civil","smp_informed","smn_manip","smn_divided"),
  Mean = c(
    mean(B$smp_civil, na.rm=TRUE),
    mean(B$smp_informed,na.rm=TRUE),
    mean(B$smn_manip,na.rm=TRUE),
    mean(B$smn_divided,na.rm=TRUE)
  )
)

stargazer(mean_BAD_impacts, # input the object "as.data.frame" for descriptives
          summary=FALSE,
          type="latex",
          rownames=FALSE,
          header = FALSE,
          title = "Social Media's Specific Impacts for Respondents Believing Social Media is Bad for Democracy",
          digits = 2)

```

According to the values in Tables 9 and 10, respondents believing social media is good for democracy think social media's ability to manipulate is the strongest impact and much less of an impact on informing people on current events going on in the country. We do, however, see a slight difference for respondents who believe social media is bad for democracy - the means for the "good" impacts (`smp_civil` and `smp_informed`) are larger than the means for the "bad" impacts, supporting the idea that those who believe social media is bad for democracy think social media's "bad" abilities to manipulate and divide is stronger than its "good" abilities to inform and promote civil discourse. In general, the means for all four variables are somewhat similar, suggesting that respondents' belief in whether social media is overall good or bad for democracy is not largely based on the defined "good" and "bad" specific impacts. (*See Tables 9 and 10*)

We are also interested in observing possible differences in satisfaction and belief that social media is good or bad for democracy by race, political ideology, age category, and whether the respondents post social and political issues on social media.

The mean satisfaction by race for respondents believing social media is good for democracy are overall similar (mostly not too satisfied or only somewhat satisfied with U.S. democracy) and slightly higher than the satisfaction of respondents believing social media is bad. Those who are mixed race and believe social media is bad for democracy have the lowest satisfaction (mean of 1.98) with U.S. democracy.

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
B_noNA <- na.omit(subset(socDem, socDem$social_demo == 0))
G_noNA <- na.omit(subset(socDem, socDem$social_demo == 1))

options(xtable.comment = FALSE)

#ATEs for race
race.good <- G_noNA %>%
  group_by(F_RACECMB) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(race.good) <- c("Race", "Mean Satisfaction")
print(xtable(race.good, caption="Respondents Believing Social Media is Good
             and Mean Satisfaction by Race"), include.rownames=FALSE, caption.placement = 'top', 
      table.placement="!h", type='latex')

race.bad <- B_noNA %>%
  group_by(F_RACECMB) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(race.bad) <- c("Race", "Mean Satisfaction")
print(xtable(race.bad, caption="Respondents Believing Social Media is Bad
             and Mean Satisfaction by Race"), include.rownames=FALSE, caption.placement = 'top',
      table.placement="!h", type='latex')
```
\newpage

Mean satisfaction still appears to be around 2 ("Not too satisfied") and are slightly lower for those who believe social media is more of a bad thing for democracy. Conservatives believing that social media is bad for democracy have the lowest satisfaction with U.S. democracy, with a mean of 1.86 (*See Table 14*).

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
#ATEs of ideology
ideology.good <- G_noNA %>%
  group_by(ideology) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(ideology.good) <- c("Ideology", "Mean Satisfaction")
print(xtable(ideology.good, caption="Respondents Believing Social Media is Good
             and Mean Satisfaction by Ideology"), 
      include.rownames=FALSE, caption.placement = 'top', table.placement = "!h", type='latex')

ideology.bad <- B_noNA %>%
  group_by(ideology) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(ideology.bad) <- c("Ideology", "Mean Satisfaction")
print(xtable(ideology.bad, caption="Respondents Believing Social Media is Bad
             and Mean Satisfaction by Ideology"), 
      include.rownames=FALSE, caption.placement = 'top', table.placement = "!h", type='latex')
```
When comparing age categories and mean satisfaction with democracy, the values remain around 2 ("Not too satisfied"), and means are slightly higher for those in the middle two age categories (ages 30 to 64). Also, the majority of respondents are in those age categories, as we observed previously. We also see that mean satisfaction is again slightly lower for those who think social media is more of a bad thing for democracy.

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
#ATEs for age
age.good <- G_noNA %>%
  group_by(F_AGECAT) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(age.good) <- c("Age Category", "Mean Satisfaction")
print(xtable(age.good, caption="Respondents Believing Social Media is Good
             and Mean Satisfaction by Age Category"), 
      include.rownames=FALSE, caption.placement = 'top', table.placement = "!h", type='latex')

age.bad <- B_noNA %>%
  group_by(F_AGECAT) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(age.bad) <- c("Age Category", "Mean Satisfaction")
print(xtable(age.bad, caption="Respondents Believing Social Media is Bad
             and Mean Satisfaction by Age Category"), 
      include.rownames=FALSE, caption.placement = 'top', table.placement = "!h", type='latex')

```
The difference in mean satisfaction for those believing social media is good for democracy compared to those believing social media is bad for democracy seems to be slightly more pronounced when we examine how frequently a respondent posts social or political issues on social media. Those that post often in both the good and bad groups have the lowest satisfaction with U.S. democracy, and those who post often *and* believe social media is bad for democracy have the lowest satisfaction (1.72) overall.

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
#ATEs for Posting
posting.good <- G_noNA %>%
  group_by(POST_ISSUES_W105) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(posting.good) <- c("Frequency of Posting",
                         "Mean Satisfaction")
print(xtable(posting.good, caption="Respondents Believing Social Media is Good
             and Mean Satisfaction by How Often They Post Issues"), 
      include.rownames=FALSE, caption.placement = 'top', table.placement = "!h", type='latex')

posting.bad <- B_noNA %>%
  group_by(POST_ISSUES_W105) %>%
  summarise_at(vars(satisfied),
               list(mean_satisfied = mean))

names(posting.bad) <- c("Frequency of Posting",
                    "Mean Satisfaction")
print(xtable(posting.bad, caption="Respondents Believing Social Media is Bad
             and Mean Satisfaction by How Often They Post Issues"), 
      include.rownames=FALSE, caption.placement = 'top', table.placement = "!h", type='latex')
```


# Statistical Analysis

## Linear Regression Analysis

We will run a linear regression analysis of believing whether social media is good or bad for democracy on satisfaction with U.S. democracy.

For our first model, we do not include race, ideology, or other variables in the regression. The p-value for this model is well below the 5% threshold, so the null hypothesis can be rejected and the belief of social media's impact on democracy does have an effect on satisfaction with democracy. The $\beta_1$ coefficient in the model (0.3245) suggests a positive correlation between our two variables, meaning those who believe social media has more of a good effect on democracy have slightly higher satisfaction ratings of U.S. democracy.

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
m1 <- lm(satisfied ~ social_demo, data = usesm_socDem)
m2 <- lm(satisfied ~ social_demo + factor(race), data = usesm_socDem)
m3 <- lm(satisfied ~ social_demo + factor(race) + factor(agecat), data = usesm_socDem)
m4 <- lm(satisfied ~ social_demo + factor(race) + factor(agecat) + factor(ideology) + factor(posting), data = usesm_socDem)
m5 <- lm(satisfied ~ social_demo + factor(race) + factor(agecat) + factor(ideology) + factor(smp_informed) + factor(smp_civil) + factor(smn_divided) + factor(smn_manip), data = usesm_socDem)
#summary(m1)

print(xtable(summary(m1), caption = 'Model 1'), caption.placement = 'top', table.placement = "!h", type='latex')


```

We then add in the other factors, beginning with race, then followed by age category, ideology, and how often the respondent posts social or political issues on social media.

When we include race as a factor, the $\beta_1$ coefficient decreases slightly to 0.2933 and still positive. The null hypothesis can still be rejected since the p-value remains below the 5% threshold.

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
print(xtable(summary(m2), caption = 'Model 2'), caption.placement = 'top', table.placement = "!h", type='latex')
```

\newpage

Next, we add ideology and age category as factors in our third model. The p-value remains statistically significant, and the null hypothesis can be rejected. The $\beta_1$ coefficient is now 0.2980, which is slightly higher than the coefficient in the previous model.

```{r, echo=FALSE, error=FALSE, warning=FALSE,  results='asis'}
print(xtable(summary(m3), caption = 'Model 3'), caption.placement = 'top', table.placement = "!h", type='latex')
```

Now we also include `posting` as a factor, since we did observe a larger difference in mean satisfaction for respondents who said they posted social and political issues on social media often.

The $\beta_1$ coefficient decreases more (down to 0.2803), but the p-value is still very small, so we can reject the null hypothesis. 

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
print(xtable(summary(m4), caption = 'Model 4'), caption.placement = 'top', table.placement = "!h", type='latex')
```

\newpage

Finally, we include all the specific possibly "good" or "bad" impacts of social media in our last model.

Similarly, the $\beta_1$ coefficient decreased, but the p-value for this model is still well below 5%, so we can reject the null hypothesis.

```{r, echo=FALSE, error=FALSE, warning=FALSE, results='asis'}
print(xtable(summary(m5), caption = 'Model 4'), caption.placement = 'top', table.placement = "!h", type='latex')
```

## Results Discussion

Factoring for all the other variables we selected that may be possible confounders, we were able to reject the null hypothesis in favor of the alternative hypothesis for all our linear regression models since the p-value remained very small (all were on the order of $10^{-12}$ or smaller). Our alternative hypothesis was that American adult’s belief that social media is “bad” or “good” for democracy does affect satisfaction/dissatisfaction with democracy, which our evidence supports.

For future research, we could include additional factors such as annual income and gender, which we did not find in the dataset we used. There could have been more datasets that the Pew Research Center report about social media's impact on democracy globally used that we did not include. We wonder if respondents with low incomes are less satisfied with U.S. democracy and whether income level is a stronger reason than perceptions of social media on satisfaction with U.S. democracy. We would also be interested in whether there is a difference between genders. For instance, we might want to determine whether women more likely to be less satisfied with the U.S. democracy and believe social media is bad for democracy.


# Conclusion

In this report, we narrowed the original ATP Wave 105 dataset to 11 variables of interest and omitted missing values. We noticed that most respondents were between 30 and 64 years of age, white, use social media, have a moderate ideology, and rarely post about social and political issues on social media. We also observed various distributions of percentages of respondents in different satisfaction categories and saw that the distributions for those who use social media had a similar distribution to that of all respondents in the dataset. However, the distributions were different when we compared those believing social media was good (more likely to be somewhat satisfied) to those believe social media was bad for democracy (more likely to be not too satisfied), which supported our causal assumption of a positive correlation. We then compared the mean satisfactions for respondents believing social media is good and respondents believing social media is bad for democracy by the variables for race, ideology, age category, and posting habits, and found a consistent trend for all for variables that those who believed social media was bad for democracy had slightly lower satisfaction means.

We ran several linear regression models, adding more factors each time, and observed that the p-value was consistently well below the 5% threshold for statistical significance for all our models, allowing us to reject the null hypothesis. The $\beta_1$ was always positive, meaning those who believe social media is more of a good thing for democracy have slightly higher satisfaction with U.S. democracy than those who believe social media is more of a bad thing for democracy, which aligns with our causal assumption and supports our alternative hypothesis. Thus, based on the evidence, a positive correlation does appear to exist between a respondent's perception of social media's impact on democracy and satisfaction with U.S. democracy. In future studies, additional effects of variables such as income and gender would be interesting to observe.

# Citations and Relevant Literature

Dixon, Stacy Jo. Social Media Usage in the United States - Statistics & Facts, Statista, 18 Dec. 2023, 
https://www.statista.com/topics/3196/social-media-usage-in-the-united-states/#topicOverview. 

Wike, Richard, et al. “Social Media Seen as Mostly Good for Democracy across Many Nations, but U.S. Is 
a Major Outlier.” Pew Research Center’s Global Attitudes Project, Pew Research Center, 6 Dec. 2022, https://www.pewresearch.org/global/2022/12/06/social-media-seen-as-mostly-good-for-democracy-across-many-nations-but-u-s-is-a-major-outlier/. 
 
## Dataset
American Trends Panel Wave 105
https://www.pewresearch.org/global/dataset/american-trends-panel-wave-105/ 






